Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-01-25 manu-draft.qmd
add costa paper and hypothesis explanation, fixes #3

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years. Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

@@

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024]. Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017]. In southwest Florida, the most common ecological example is the expansion of mangroves upland as porewater salinity and water levels have gradually increased over the past few decades [@borchert2018]. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Sea level rise is further altering salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity, especially in the upper bay segments, suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.

>>>>

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years. Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community, despite information in the primary literature suggesting the contrary for salinity [e.g., @costa2023]. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

@@

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@costa2023; @alarcon2024]. Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017]. In southwest Florida, the most common ecological example is the expansion of mangroves upland as porewater salinity and water levels have gradually increased over the past few decades [@borchert2018]. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Sea level rise is further altering salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity, especially in the upper bay segments, suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.
